---
title: "Acoustic and Kinematic Predictors of Intelligibility and Articulatory Precision in Parkinson’s Disease"
---


```{r}
# Starting with a clean environment
rm(list = ls())
```

# Packages
```{r}
library(tidyverse) # install.packages('tidyverse')
library(emmeans)
library(robustlmm) # install.packages("robustlmm")
library(ggpubr) # install.packages('ggpubr')
library(plyr) # install.packages('plyr')
library(beepr) # install.packages('beepr')
library(plotrix) # install.packages('plotrix')
```

# Load Data
```{r}
speakerList <- rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/Speaker%20List_Clean.csv")
```

## Listener Descriptions
```{r}
listenerRatings <- rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/ListenerData/ListenerRatings_allRatings.csv")

ratingCount <- listenerRatings %>%
  dplyr::group_by(StudyID, ratingType) %>%
  dplyr::summarise(n = NROW(Rating)) %>%
  dplyr::ungroup()

ratingCount %>%
  dplyr::summarise(mean = mean(n),
                   sd = sd(n),
                   min = min(n),
                   max = max(n))

NROW(unique(listenerRatings$ListenerID))

listenerDemo <- rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/ListenerData/ListenerDemographics.csv")
NROW(unique(listenerDemo$ListenerID))

# Age and N
listenerDemo %>%
  dplyr::mutate(duration = `Duration (in seconds)`/60) %>%
  #dplyr::group_by(gender) %>%
  dplyr::summarise(age_m = mean(age, na.rm = T),
                   age_sd = sd(age, na.rm = T),
                   min = base::min(age, na.rm = T),
                   max = base::max(age, na.rm = T),
                   N = NROW(age),
                   duration_M = base::mean(duration, na.rm = T),
                   duration_sd = sd(duration, na.rm = T))

listenerDemo %>%
  dplyr::mutate(duration = `Duration (in seconds)`/60) %>%
  dplyr::group_by(gender) %>%
  dplyr::summarise(N = NROW(age))

listenerDemo %>%
  dplyr::mutate(duration = `Duration (in seconds)`/60) %>%
  dplyr::group_by(race) %>%
  dplyr::summarise(N = NROW(race))

listenerDemo %>%
  dplyr::mutate(duration = `Duration (in seconds)`/60) %>%
  dplyr::group_by(ethnicity) %>%
  dplyr::summarise(N = NROW(ethnicity))
```

# APA reporting function
```{r}

apa_lmResults <- function(model, fixedEffect) {
  
coefficients <- summary(model)[["coefficients"]] %>%
  as.data.frame() %>%
  dplyr::rename(
    stdError = `Std. Error`,
    t = `t value`,
    p = `Pr(>|t|)`) %>%
  dplyr::mutate(
    Estimate = round(Estimate, digits = 2),
    #df = round(df, digits = 2),
    t = round(t, digits = 2),
    p = round(p, digits = 3),
  ) %>%
  dplyr::filter(rownames(.) == fixedEffect)

if(coefficients$p < .001) {
  p <- "p<.001"
} else {
  p <- paste0("p=",as.numeric(coefficients$p))
}

df <- as.numeric(model$df.residual)
t <- paste0("t(",df,")=", as.numeric(coefficients$t))

CI_intervals <- stats::confint(model) %>%
  as.data.frame() %>%
  dplyr::mutate(
    ll = round(`2.5 %`, digits = 2),
    ul = round(`97.5 %`, digits = 2),
                ) %>%
  dplyr::select(ll, ul) %>%
  dplyr::filter(rownames(.) == fixedEffect)

CI <- base::paste0("95% CI [",
                  as.numeric(CI_intervals$ll),
                  ", ",
                  as.numeric(CI_intervals$ul),
                  "]")

formula <- paste(t, p, CI, sep = ", ")

return(formula)

rm(coefficients, t, p, CI_intervals, CI)
}

apa_lmeResults <- function(model, fixedEffect) {
  
coefficients <- summary(model)[["coefficients"]] %>%
  as.data.frame() %>%
  dplyr::rename(
    stdError = `Std. Error`,
    t = `t value`,
    p = `Pr(>|t|)`) %>%
  dplyr::mutate(
    Estimate = round(Estimate, digits = 2),
    df = round(df, digits = 2),
    t = round(t, digits = 2),
    p = round(p, digits = 3),
  ) %>%
  dplyr::filter(rownames(.) == fixedEffect)

if(coefficients$p < .001) {
  p <- "p<.001"
} else {
  p <- paste0("p=",as.numeric(coefficients$p))
}

t <- paste0("t(",as.numeric(coefficients$df),")=", as.numeric(coefficients$t))

CI_intervals <- stats::confint(model) %>%
  as.data.frame() %>%
  dplyr::mutate(
    ll = round(`2.5 %`, digits = 2),
    ul = round(`97.5 %`, digits = 2),
                ) %>%
  dplyr::select(ll, ul) %>%
  dplyr::filter(rownames(.) == fixedEffect)

CI <- base::paste0("95% CI [",
                  as.numeric(CI_intervals$ll),
                  ", ",
                  as.numeric(CI_intervals$ul),
                  "]")

formula <- paste(t, p, CI, sep = ", ")

return(formula)

rm(coefficients, t, p, CI_intervals, CI)
}

apa_levelDescriptives <- function(data,level, outcome) {

  
descriptiveData <- modelData %>%
  dplyr::select(Group, Sex, sym(outcome)) %>%
  dplyr::rename(outcome = 3)

group <- descriptiveData %>%
  dplyr::group_by(Group) %>%
  dplyr::summarize(M = mean(outcome, na.rm = T),
                SD = sd(outcome, na.rm = T),
                .groups = 'drop') %>%
  dplyr::mutate(M = round(M, digits = 2),
                SD = round(SD, digits = 2)) %>%
  dplyr::rename(Grouping = 1) %>%
  as.data.frame()

sex <- descriptiveData %>%
  dplyr::group_by(Sex) %>%
  dplyr::summarize(M = mean(outcome, na.rm = T),
                SD = sd(outcome, na.rm = T),
                .groups = 'drop') %>%
  dplyr::mutate(M = round(M, digits = 2),
                SD = round(SD, digits = 2)) %>%
  dplyr::rename(Grouping = 1) %>%
  as.data.frame()

groupSex <- descriptiveData %>%
  dplyr::group_by(Group, Sex) %>%
  dplyr::summarize(M = mean(outcome, na.rm = T),
                SD = sd(outcome, na.rm = T),
                .groups = 'drop') %>%
  dplyr::ungroup() %>%
  dplyr::mutate(M = round(M, digits = 2),
                SD = round(SD, digits = 2),
                Grouping = paste(Group, Sex, sep = "_")) %>%
  dplyr::select(Grouping, M, SD) %>%
  as.data.frame()

descriptives <- rbind(group,
                      sex,
                      groupSex) %>%
  dplyr::filter(Grouping == level)
  
  output <- paste0(level,": (M=",descriptives$M,", SD=",descriptives$SD,")")
  
  return(output)
  
  rm(descriptives, group, sex, groupSex)
  
}


apa_emmsPairs <- function(emmsPair, sex, Condition) {
  summary_table <- summary(emmsPair, infer = TRUE) %>%
    dplyr::filter(Sex == sex) %>%
    dplyr::filter(condition == condition)
  
if(summary_table$p.value < .001) {
  p <- "p < .001"
} else {
    p <- paste0("p=",
                round(summary_table$p.value, digits = 3))
  }
  
  if("t.ratio" %in% colnames(summary_table)) {
  output <- paste(
    sex,
    ", HC - PD: (",
    "t(",
    round(summary_table$df, digits = 2),
    ")=",
    as.numeric(round(summary_table$t.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  
  if("z.ratio" %in% colnames(summary_table)) {
  output <- paste(
    sex,
    ", HC - PD: (",
    "z = (",
    as.numeric(round(summary_table$z.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  return(output)
}

apa_s1_emmPairs <- function(emmsPair, sex, Condition) {
  summary_table <- summary(emmsPair, infer = TRUE) %>%
    dplyr::filter(Sex == sex) %>%
    dplyr::filter(condition == Condition)
  
  if(summary_table$p.value < .001) {
  p <- "p<.001"
} else {
    p <- paste0("p=",
                round(summary_table$p.value, digits = 3))
  }
  
  if("t.ratio" %in% colnames(summary_table)) {
  output <- paste(
    sex,
    ", HC - PD: (",
    "t(",
    round(summary_table$df, digits = 2),
    ")=",
    as.numeric(round(summary_table$t.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  
  if("z.ratio" %in% colnames(summary_table)) {
  output <- paste(
    sex,
    ", HC - PD: (",
    "z=",
    as.numeric(round(summary_table$z.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  return(output)
}

apa_s2_emmPairs <- function(emmsPair, Sex_Input, Group_Input, Contrast_Input) {
  summary_table <- summary(emmsPair, infer = TRUE) %>%
    dplyr::filter(Sex == Sex_Input) %>%
    dplyr::filter(Group == Group_Input) %>%
    dplyr::filter(contrast == Contrast_Input)
  
  if(summary_table$p.value < .001) {
  p <- "p<.001"
} else {
    p <- paste0("p=",
                round(summary_table$p.value, digits = 3))
  }
  
  if("t.ratio" %in% colnames(summary_table)) {
  output <- paste(
    Sex_Input,
    ", ",
    Group_Input,
    ", ",
    Contrast_Input,
    ": (",
    "t(",
    round(summary_table$df, digits = 2),
    ")=",
    as.numeric(round(summary_table$t.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  
  if("z.ratio" %in% colnames(summary_table)) {
  output <- paste(
    Sex_Input,
    ", ",
    Group_Input,
    ", ",
    Contrast_Input,
    ": (",
    "z=",
    as.numeric(round(summary_table$z.ratio, 2)),
    ", ",
    p,
    ")",
    sep = "")
  }
  
  return(output)
}



```


# RQ1 - Correlations
1.	What is the nature and extent of the relationships between perceptual, acoustic, and kinematic measures across conversational, less clear, and more clear speaking conditions?
    a.	What is the relationship between intelligibility and articulatory precision ratings?
    b.	What is the relationship between acoustic measures and their underlying theoretical kinematic measures (i.e., acoustic VSA vs. kinematic VSA, acoustic distance vs. kinematic distance, F2 slope vs. kinematic speed)?
```{r}
vsaMeasures <- rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/CollatedData/TargetMeasures_vsaMeasures.csv") %>%
  dplyr::mutate(Sex = factor(Sex, levels = c("M","F")),
                aVSA = aVSA/1000)

aiMeasures <- rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/CollatedData/TargetMeasures_aiMeasures.csv") %>%
  dplyr::mutate(Sex = factor(Sex, levels = c("M","F")))
```


### 1A - Int & AP
What is the relationship between intelligibility and articulatory precision ratings?
```{r}

# VSA Data
stats::cor.test(x = vsaMeasures$Int_M,
                y = vsaMeasures$AP_M,
                method = "pearson")

# Diphthong Data
stats::cor.test(x = aiMeasures$Int_M,
                y = aiMeasures$AP_M,
                method = "pearson")

```

## Correlations
```{r}

# VSA Data
corrr::correlate(vsaMeasures %>% dplyr::select(Int_M, AP_M, aVSA, kVSA))

# Diphthong Data
corrr::correlate(aiMeasures %>% 
                               dplyr::select(Int_M,
                                             AP_M,
                                             acoDistance,
                                             F2_Slope,
                                             kinDistance,
                                             kinSpeed),
                             method = "pearson")


stats::cor.test(x = vsaMeasures$aVSA,
                y = vsaMeasures$kVSA,
                method = "pearson")
```
# RQ2 - Models
2.	What is the relationship between the selected acoustic/kinematic measures and perceptual ratings of speech intelligibility and articulatory precision across the three speaking conditions?
## 1. Point-Based Measures
```{r}
vsaMeasures <-
  rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/CollatedData/TargetMeasures_vsaMeasures.csv") %>%
  dplyr::mutate(Sex = factor(Sex, levels = c("M", "F")),
                aVSA = aVSA / 1000)

modelData <- vsaMeasures %>%
  dplyr::rename(SpeakerID = StudyID)

# Taking out the trash
rm(vsaMeasures)
```

### Intelligibility
M1 - Comprehensive Model
```{r}
Int_PBS_m1 <- robustlmm::rlmer(Int_M ~
                                 aVSA * Sex +
                                 #kVSA * Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(Int_PBS_m1)
```
We can see that only the aVSA, Sex, and aVSA x Sex predictors were significant. So we will only retain those in the most parsimonious model.

M2
```{r}
Int_PBS_m2 <- robustlmm::rlmer(Int_M ~
                                 aVSA * Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(Int_PBS_m2)
```

Final Model
```{r}
Int_PBS <- Int_PBS_m2

summary(Int_PBS)
sjPlot::tab_model(Int_PBS)

# Saving the model
base::saveRDS(object = Int_PBS,
              file = "Models/Int_PBS.rds")

# Taking out the trash
rm(list = ls(pattern = "^Int_m"))
rm(list = ls(pattern = "^emms_"))
```

### Articulatory Precision
M1 - Comprehensive Model
```{r}
AP_PBS_m1 <- robustlmm::rlmer(AP_M ~
                                aVSA * Sex +
                                kVSA * Sex +
                                (1 | SpeakerID),
                              data = modelData)

sjPlot::tab_model(AP_PBS_m1)
```
We can see that only the aVSA, Sex, and aVSA x Sex predictors were significant. So we will only retain those in the most parsimonious model.

M2
```{r}
AP_PBS_m2 <- robustlmm::rlmer(AP_M ~
                                 aVSA * Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(AP_PBS_m2)
```

Final Model
```{r}
AP_PBS <- AP_PBS_m2

summary(AP_PBS)
sjPlot::tab_model(AP_PBS)

# Saving the model
base::saveRDS(object = AP_PBS,
              file = "Models/AP_PBS.rds")

# Taking out the trash
rm(list = ls(pattern = "^AP_m"))
rm(list = ls(pattern = "^emms_"))
```

### Saving Models
Saving the models for the table
```{r}
base::saveRDS(Int_PBS_m1, file = "Models/Int_PBS_m1.rds")
base::saveRDS(AP_PBS_m1, file = "Models/AP_PBS_m1.rds")
```

## 2. Transition-Based Measures
```{r}
aiMeasures <-
  rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/CollatedData/TargetMeasures_aiMeasures.csv") %>%
  dplyr::mutate(Sex = factor(Sex, levels = c("M", "F")))

modelData <- aiMeasures %>%
  dplyr::rename(SpeakerID = StudyID)

# Taking out the trash
rm(aiMeasures)
```

### Intelligibility
M1 - Comprehensive Model
```{r}
Int_TBS_m1 <- robustlmm::rlmer(Int_M ~
                                 acoDistance * Sex +
                                 kinDistance * Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(Int_TBS_m1)
```
We can see that only the kinDistance, and Sex were significant. So we will only retain those in the most parsimonious model.

M2
```{r}
Int_TBS_m2 <- robustlmm::rlmer(Int_M ~
                                 kinDistance +
                                 Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(Int_TBS_m2)
```

Final Model
```{r}
Int_TBS <- Int_TBS_m2

summary(Int_TBS)
sjPlot::tab_model(Int_TBS)

# Saving the model
base::saveRDS(object = Int_TBS,
              file = "Models/Int_TBS.rds")

# Taking out the trash
rm(list = ls(pattern = "^Int_m"))
rm(list = ls(pattern = "^emms_"))
```
### Articulatory Precision
M1 - Comprehensive Model
```{r}
AP_TBS_m1 <- robustlmm::rlmer(AP_M ~
                                 acoDistance * Sex +
                                 kinDistance * Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(AP_TBS_m1)
```
We can see that only the kinDistance, and Sex were significant. So we will only retain those in the most parsimonious model.

M2 - Comprehensive Model
```{r}
AP_TBS_m2 <- robustlmm::rlmer(AP_M ~
                                 kinDistance +
                                 Sex +
                                 (1 | SpeakerID),
                               data = modelData)

sjPlot::tab_model(AP_TBS_m2)
```

Final Model
```{r}
AP_TBS <- AP_TBS_m2

summary(AP_TBS)
sjPlot::tab_model(AP_TBS)

# Saving the model
base::saveRDS(object = AP_TBS,
              file = "Models/AP_TBS.rds")

# Taking out the trash
rm(list = ls(pattern = "^AP_m"))
rm(list = ls(pattern = "^emms_"))
```
### Saving Models
Saving the models for the table
```{r}
base::saveRDS(Int_TBS, file = "Models/Int_TBS_m1.rds")
base::saveRDS(AP_TBS_m1, file = "Models/AP_TBS_m1.rds")
```

## 3. Spatiotemporal Measures
```{r}
aiMeasures <-
  rio::import(file = "https://raw.githubusercontent.com/AustinRThompson/multidomain-vowelArtic-PD/main/Data/PreppedData/CollatedData/TargetMeasures_aiMeasures.csv") %>%
  dplyr::mutate(Sex = factor(Sex, levels = c("M", "F")))

modelData <- aiMeasures %>%
  dplyr::rename(SpeakerID = StudyID)

# Taking out the trash
rm(aiMeasures)
```

### Intelligibility
M1
```{r}
# Building the model
Int_SpaTemp_m1 <- robustlmm::rlmer(Int_M ~
                                     F2_Slope * Sex +
                                     kinSpeed * Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(Int_SpaTemp_m1)
```
Only sex was a significant predictor of intelligibility.

M2
```{r}
# Building the model
Int_SpaTemp_m2 <- robustlmm::rlmer(Int_M ~
                                     F2_Slope +
                                     kinSpeed +
                                     Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(Int_SpaTemp_m2)
```

M3
```{r}
# Building the model
Int_SpaTemp_m3 <- robustlmm::rlmer(Int_M ~
                                     Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(Int_SpaTemp_m3)
```

Final Model
```{r}
Int_SpaTemp <- Int_SpaTemp_m3

summary(Int_SpaTemp)
sjPlot::tab_model(Int_SpaTemp)

# Saving the model
base::saveRDS(object = Int_SpaTemp,
              file = "Models/Int_SpaTemp.rds")

# Taking out the trash
rm(list = ls(pattern = "^Int_m"))
rm(list = ls(pattern = "^emms_"))
```

### Articulatory Precision
M1
```{r}
# Building the model
AP_SpaTemp_m1 <- robustlmm::rlmer(AP_M ~
                                     F2_Slope * Sex +
                                     kinSpeed * Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(AP_SpaTemp_m1)
```
Only sex was a significant predictor of articulatory precision

M2
```{r}
# Building the model
AP_SpaTemp_m2 <- robustlmm::rlmer(AP_M ~
                                     F2_Slope +
                                     kinSpeed +
                                     Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(AP_SpaTemp_m2)
```
Removing the sex interactions revealed that F2 slope appears to be a meaningful predictor of articulatory precision. Let's continue to remove non significant variables to obtain the most parsimonious model.

M3
```{r}
# Building the model
AP_SpaTemp_m3 <- robustlmm::rlmer(AP_M ~
                                     F2_Slope +
                                     Sex +
                                     (1 | SpeakerID),
                                   data = modelData)
# Model summary
sjPlot::tab_model(AP_SpaTemp_m3)
```

Final Model
```{r}
AP_SpaTemp <- AP_SpaTemp_m3

summary(AP_SpaTemp)
sjPlot::tab_model(AP_SpaTemp)

# Saving the model
base::saveRDS(object = AP_SpaTemp,
              file = "Models/AP_SpaTemp.rds")

# Taking out the trash
rm(list = ls(pattern = "^AP_m"))
rm(list = ls(pattern = "^emms_"))
```

### Saving Models
Saving the models for the table
```{r}
base::saveRDS(Int_SpaTemp_m1, file = "Models/Int_SpaTemp_m1.rds")
base::saveRDS(Int_SpaTemp_m2, file = "Models/Int_SpaTemp_m2.rds")
base::saveRDS(AP_SpaTemp_m1, file = "Models/AP_SpaTemp_m1.rds")
base::saveRDS(AP_SpaTemp_m2, file = "Models/AP_SpaTemp_m2.rds")
```
